---
title: "Data Visualization with the Tidyverse in R"
editor: visual
author: "Jason Laird"
date: "05/29/2024"
format: 
  html:
    toc: true
    toc-location: left
    theme: 
      - flatly
    linkcolor: '#555580'
---

## Introduction To ggplot2

`ggplot2` is by far one of the premier plotting packages offered through R. Here we will walk through the basics of how to make a publication ready plot!

## Load Libraries

```{r}
library(tidyverse)
library(ggpubr)
```

## Load Data

```{r}
brainspan <- readr::read_csv("../results/brainspan.csv")
```

## Creating Different Types of Plots

### Scatter Plot

```{r}
brainspan %>% 
  ggplot(
    aes(
      x=GAPDH,
      y=ACTB,
      color=gender
    )
  )+
  geom_point()
```

### Bar Plots

```{r}
brainspan %>% 
  tabyl(structure_acronym,gender) %>% 
  pivot_longer(!structure_acronym,names_to="gender",values_to="value") %>% 
  ggplot(
    aes(
      x=reorder(structure_acronym,-value),
      y=value,
      fill=gender
    )
  )+
  geom_bar(stat="identity")+
  rotate_x_text(angle=60)
```

### Histograms (Density Plot)

```{r}
brainspan %>% 
  ggplot(aes(
    x=GAPDH
  ))+
  geom_density(fill="violetred4")
```

### Violin Plots

```{r}
brainspan %>% 
  ggplot(aes(
    x=gender,
    y=GAPDH,
    fill=gender
  ))+
  geom_violin()
```

## Customizing Plots

### Adjusting the Axis

```{r}
brainspan %>% 
  ggplot(aes(
    x=gender,
    y=GAPDH,
    fill=gender
  ))+
  geom_violin()+
  scale_y_log10()
```

### Colors

```{r}
brainspan %>% 
  ggplot(
    aes(
      x=GAPDH,
      y=ACTB,
      color=gender
    )
  )+
  geom_point()+
  scale_color_manual(values=c("magenta4","midnightblue"))
```

### Shapes

```{r}
brainspan %>% 
  ggplot(
    aes(
      x=GAPDH,
      y=ACTB,
      color=gender,
      shape = gender
    )
  )+
  geom_point()+
  scale_color_manual(values=c("magenta4","midnightblue"))
```

### Sizes

```{r}
brainspan %>% 
  ggplot(
    aes(
      x=GAPDH,
      y=ACTB,
      color=gender,
      shape = gender,
      size = GAPDH
    )
  )+
  geom_point()+
  scale_color_manual(values=c("magenta4","midnightblue"))
```

### Adjusting the Text

```{r}
brainspan %>% 
  ggplot(
    aes(
      x=GAPDH,
      y=ACTB,
      color=gender,
      shape = gender,
      size = GAPDH
    )
  )+
  geom_point()+
  scale_color_manual(values=c("magenta4","midnightblue"))+
  labs(
    x="GAPDH",
    y="ACTB",
    size="GAPDH",
    color="Gender",
    shape = "Gender",
    title="Scatterplot GAPDH v. ACTB",
    subtitle = "males v. females",
    caption = "For educational purposes only"
  )
```

### Faceting

```{r}
brainspan %>% 
  ggplot(
    aes(
      x=GAPDH,
      y=ACTB,
      color=gender,
      shape = gender,
      size = GAPDH
    )
  )+
  geom_point()+
  scale_color_manual(values=c("magenta4","midnightblue"))+
  labs(
    x="GAPDH",
    y="ACTB",
    size="GAPDH",
    color="Gender",
    shape = "Gender",
    title="Scatterplot GAPDH v. ACTB",
    subtitle = "males v. females",
    caption = "For educational purposes only"
  )+
  facet_grid(.~gender)
```

### General Themes

```{r}
brainspan %>% 
  ggplot(
    aes(
      x=GAPDH,
      y=ACTB,
      color=gender,
      shape = gender,
      size = GAPDH
    )
  )+
  geom_point()+
  scale_color_manual(values=c("magenta4","midnightblue"))+
  labs(
    x="GAPDH",
    y="ACTB",
    size="GAPDH",
    color="Gender",
    shape = "Gender",
    title="Scatterplot GAPDH v. ACTB",
    subtitle = "males v. females",
    caption = "For educational purposes only"
  )+
  facet_grid(.~gender)+
  theme_pubr(legend="right")
```

### Customize the Theme

```{r}
brainspan %>% 
  ggplot(
    aes(
      x=GAPDH,
      y=ACTB,
      color=gender,
      shape = gender,
      size = GAPDH
    )
  )+
  geom_point()+
  scale_color_manual(values=c("magenta4","midnightblue"))+
  labs(
    x="GAPDH",
    y="ACTB",
    size=expression("Log"[10]*gamma),
    color="Gender",
    shape = "Gender",
    title= "Scatterplot GAPDH v. ACTB",
    subtitle = "males v. females",
    caption = "For educational purposes only"
  )+
  facet_grid(.~gender)+
  theme_pubr(legend="right")+
  theme(plot.title = element_text(face="bold.italic"))
```

## Saving Plots

```{r}
ggsave(
  filename = "../results/finished_scatterplot.png",
  plot = last_plot()
)
```
